---
title: "Finding Discrepancy in Data-Files using Tidyverse"
author: "Monika Baloda"
date: "2023-04-07"
output:
  html_document: default
  pdf_document: default
---



***The Problem***
There are two dataframe recording observations some fields on same IDs, some of them are overlaping in both the datasets. 
Each data frame has 100k IDs and 11 fields.  
Not all of the IDs or fields match between both files; 70k IDs match and 7 fields + ID match between both files. 
Of the 70k IDs that match, and the 7 fields that match, all values should be the same in both files. If they are not, then we call it discrepancy. 
We need to find the discrepancies from these two datasets and you must find them.


***Analysis***

**Step 0: loading the library**
```{r, warning=F}
library(tidyverse)
```


**Step 1: Read two files**
```{r}
setwd("C:/Users/EndUser/Desktop/UCR/Spring2023/STAT208/Challanges/Challange1")

df1=read.csv("Challenge1_Dataframe1.csv")
df2=read.csv("Challenge1_Dataframe2.csv")
```


**Step 2: Merge dataframes by ID and keep common 7 fields from each of the dataframe**

We use inner join to keep the IDs that matches in both the dataframes. 
```{r}
#merging dataframes
merged_df = inner_join(df1, df2, by = "ID")

#keeping matched fields only
merged_df_selected = select(merged_df, ID, Income..Dollars..x,Gender.x,Town.x,Region.x,
                                Internal.IP.Address.x,Customer.Age.Years.x, Encrypted.Password.x,
                            Income..Dollars..y,Gender.y,Town.y,Region.y,
                          Internal.IP.Address.y,Customer.Age.Years.y,Encrypted.Password.y)
```


**Step 3 : Comparing the values of individual fields of two dataframes**


*Comparing income from df1 to income from df2*
```{r}
# compare values in col1 and col2 and return logical vector
false_rows_income = which(merged_df_selected$Income..Dollars..x != merged_df_selected$Income..Dollars..y)

#this prints ID number of the dataset.
merged_df_selected[false_rows_income, "ID"]
```

*Comparing gender from df1 to income from df2*
```{r}
false_rows_gender = which(merged_df_selected$Gender.x != merged_df_selected$Gender.y)
merged_df_selected[false_rows_gender, "ID"]
```

```{r}
false_rows_Town = which(merged_df_selected$Town.x != merged_df_selected$Town.y)
merged_df_selected[false_rows_Town, "ID"]
```

```{r}
false_rows_Region = which(merged_df_selected$Region.x != merged_df_selected$Region.y)
merged_df_selected[false_rows_Region, "ID"]
```

```{r}
false_rows_IP= which(merged_df_selected$Internal.IP.Address != merged_df_selected$Internal.IP.Address.y)
merged_df_selected[false_rows_IP, "ID"]

```

```{r}
false_rows_age = which(merged_df_selected$Customer.Age.Years.x != merged_df_selected$Customer.Age.Years.y)
merged_df_selected[false_rows_age, "ID"]

```

```{r}
false_rows_pass = which(merged_df_selected$Encrypted.Password.x != merged_df_selected$Encrypted.Password.y)
merged_df_selected[false_rows_pass, "ID"]

```

***Final Answer***
So the IDs with discrepancies are: 


    ID       Field
    
1. 140701   Customer.Age.Years

2. 934535   Customer.Age.Years

3. 655190    Town

4. 59305     Income..Dollars.

5. 497167   Income..Dollars.

6. 860991   Region

7. 410858  Encrypted.Password

8. 985681  Encrypted.Password