---
title: "Finding Aggregate Savings of Customers by Joining Different Tables"
author: "Monika Baloda"
date: "2023-04-07"
output:
  pdf_document: default
  html_document: default
---



***The Problem***

Work in SQL databases for data files, work for corporate, data competitions like datafest / kaggle, 
then data is often stored in multiple tables. 
Analyses are unable to be completed unless you know how to join in the needed data. 

This Challenge will require all 3 data sources:

Data-1 contains ID and Savings. 

Data-2 contains ID, Tag and several other variables. 

Data-3 contains Tag and Customer tiers. 


*The Goal* is to calculate the mean Savings for all 5 customer tiers, answer example:


Customer Tier 1: XXXXXX

Customer Tier 2: YYYYYY

...

Customer Tier 5: ZZZZZZ

 
 

***Solution***

**Step 0: loading the library**
```{r, warning=F}
#library(tidyverse)
```


**Step 1: Read two files**
```{r}
setwd("C:/Users/EndUser/Desktop/UCR/Spring2023/STAT208/Challanges/Challange2")

tag=read.csv("Challenge2_CustomerTag.csv")
saving=read.csv("Challenge2_Savings.csv")
tier=read.csv("Challenge2_TagTier.csv")
```


**Step 2: 1:1 Merging dataframes : tag and savings**

```{r}
df=merge(tag, saving, by = "ID")
names(df)  #prints columns of merged dataframe. If saving column is there, we are fine
```


**Step-3 : 1:n Merging : merging df and tier with `tag`**

```{r}
df_final=merge(df, tier, by = "Tag")
dim(df_final)   #number of rows are 100000, so we are fine
names(df_final)  #if customer.Tier appears in columns, we are fine
#head(df_final)  #one can have a look at final dataframe
```


**Step-4 : Aggregating Savings based on Customer Tiers**

```{r}
agg_df = aggregate(Savings ~ Customer.Tier, data = df_final, FUN = sum)
```

***Printing the Final Answer***
```{r}
agg_df
```


